---
title: "Visualization with scCulturePredict"
author: "Niccolò Bianchi"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Visualization with scCulturePredict}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 8,
    fig.height = 6,
    warning = FALSE,
    message = FALSE
)
```

# Introduction

Effective visualization is crucial for interpreting single-cell RNA sequencing data. The `scCulturePredict` package provides a comprehensive set of visualization functions designed to help you explore your data and evaluate prediction results. This vignette demonstrates how to use these visualization tools to create informative and publication-ready figures.

## Loading Required Packages

```{r setup}
library(scCulturePredict)
library(Seurat)
library(ggplot2)
library(patchwork)
```

## Creating Example Data

For this vignette, we'll create some example data to demonstrate the visualization capabilities:

```{r example_data}
# Create a mock Seurat object
create_mock_seurat <- function() {
    # Create a small Seurat object with UMAP coordinates
    set.seed(42)
    counts <- matrix(rpois(100 * 30, 10), nrow = 100, ncol = 30)
    rownames(counts) <- paste0("gene", 1:100)
    colnames(counts) <- paste0("cell", 1:30)

    seurat <- Seurat::CreateSeuratObject(counts = counts)

    # Add UMAP coordinates and metadata
    seurat@meta.data$UMAP_1 <- rnorm(30)
    seurat@meta.data$UMAP_2 <- rnorm(30)
    seurat@meta.data$sample <- rep(c("A", "B", "C"), each = 10)
    seurat@meta.data$cluster <- rep(c(1, 2, 3), times = 10)
    seurat@meta.data$percent_mito <- runif(30, 0, 10)

    return(seurat)
}

# Create mock evaluation results
create_mock_evaluation <- function() {
    list(
        direct_accuracy = data.frame(
            sample = c("A", "B", "C"),
            correct = c(8, 7, 9),
            percent = c("80%", "70%", "90%")
        ),
        direct_table = as.table(matrix(c(8, 1, 1, 1, 7, 2, 1, 2, 7),
            nrow = 3,
            dimnames = list(c("A", "B", "C"), c("A", "B", "C"))
        )),
        svm_accuracy = data.frame(
            sample = c("A", "B", "C"),
            correct = c(9, 8, 7),
            percent = c("90%", "80%", "70%")
        ),
        svm_table = as.table(matrix(c(9, 0, 1, 1, 8, 1, 0, 2, 7),
            nrow = 3,
            dimnames = list(c("A", "B", "C"), c("A", "B", "C"))
        )),
        threshold_table = as.table(matrix(c(7, 1, 0, 2, 6, 1, 1, 2, 8),
            nrow = 3,
            dimnames = list(c("A", "B", "C"), c("A", "B", "C"))
        ))
    )
}

# Generate example data
seurat_object <- create_mock_seurat()
evaluation_results <- create_mock_evaluation()
```

# Basic Visualizations

## UMAP Visualization

The `create_umap_plot` function generates UMAP visualizations from a Seurat object:

```{r umap_basic}
# Basic UMAP plot
umap_plot <- create_umap_plot(
    seurat_object = seurat_object,
    color_by = "sample",
    title = "UMAP of Example Data"
)

print(umap_plot)
```

### Customizing UMAP Plots

You can customize various aspects of the UMAP plot:

```{r umap_custom}
# Customized UMAP plot
umap_custom <- create_umap_plot(
    seurat_object = seurat_object,
    color_by = "cluster",
    point_size = 2.5,
    title = "UMAP by Cluster",
    legend_title = "Cluster ID"
)

print(umap_custom)
```

You can also use other metadata columns for coloring:

```{r umap_continuous}
# UMAP with continuous variable
umap_continuous <- create_umap_plot(
    seurat_object = seurat_object,
    color_by = "percent_mito",
    point_size = 2,
    title = "UMAP by Mitochondrial Percentage",
    legend_title = "% Mito"
)

print(umap_continuous)
```

### Combining Multiple UMAP Plots

The plots can be combined using the `patchwork` package:

```{r umap_combined}
# Combine multiple UMAP plots
combined_plot <- umap_plot + umap_custom + umap_continuous +
    plot_layout(ncol = 2) +
    plot_annotation(title = "Multiple UMAP Visualizations")

print(combined_plot)
```

# Prediction Evaluation Visualizations

## Accuracy Plots

The `create_accuracy_plot` function visualizes prediction accuracy:

```{r accuracy}
# Direct method accuracy plot
direct_accuracy <- create_accuracy_plot(
    evaluation_results = evaluation_results,
    method = "direct",
    title = "Direct Prediction Accuracy"
)

# SVM method accuracy plot
svm_accuracy <- create_accuracy_plot(
    evaluation_results = evaluation_results,
    method = "svm",
    title = "SVM Prediction Accuracy"
)

# Display the plots side by side
direct_accuracy + svm_accuracy + plot_layout(ncol = 2)
```

### Customizing Accuracy Plots

You can customize the color palette:

```{r accuracy_custom}
# Custom color palette
custom_colors <- c("A" = "#E41A1C", "B" = "#377EB8", "C" = "#4DAF4A")

# Customized accuracy plot
custom_accuracy <- create_accuracy_plot(
    evaluation_results = evaluation_results,
    method = "direct",
    title = "Customized Accuracy Plot",
    color_palette = custom_colors
)

print(custom_accuracy)
```

## Confusion Matrix Heatmaps

The `create_confusion_heatmap` function visualizes confusion matrices:

```{r confusion}
# Direct method confusion matrix
direct_confusion <- create_confusion_heatmap(
    evaluation_results = evaluation_results,
    method = "direct",
    title = "Direct Prediction Confusion Matrix"
)

# SVM method confusion matrix
svm_confusion <- create_confusion_heatmap(
    evaluation_results = evaluation_results,
    method = "svm",
    title = "SVM Prediction Confusion Matrix"
)

# Threshold method confusion matrix
threshold_confusion <- create_confusion_heatmap(
    evaluation_results = evaluation_results,
    method = "threshold",
    title = "Threshold Prediction Confusion Matrix"
)

# Display the plots
direct_confusion
svm_confusion
threshold_confusion
```

### Customizing Confusion Heatmaps

You can customize the color gradient:

```{r confusion_custom}
# Custom color gradient
custom_gradient <- c("white", "navy")

# Customized confusion heatmap
custom_confusion <- create_confusion_heatmap(
    evaluation_results = evaluation_results,
    method = "svm",
    title = "Customized Confusion Matrix",
    color_palette = custom_gradient
)

print(custom_confusion)
```

# Evaluation Metrics Visualization

The `create_evaluation_metrics_plot` function provides detailed evaluation metrics:

```{r eval_metrics, eval=FALSE}
# For this demonstration, we need additional metrics that would come from a real evaluation
# This code is provided as an example but not evaluated in this vignette

# Create evaluation metrics plot
metrics_plot <- create_evaluation_metrics_plot(
    evaluation_results = complete_evaluation_results,
    plot_type = "metrics",
    title = "Prediction Performance Metrics"
)

print(metrics_plot)

# ROC curve
roc_plot <- create_evaluation_metrics_plot(
    evaluation_results = complete_evaluation_results,
    plot_type = "roc",
    title = "ROC Curve"
)

print(roc_plot)

# Precision-Recall curve
pr_plot <- create_evaluation_metrics_plot(
    evaluation_results = complete_evaluation_results,
    plot_type = "pr",
    title = "Precision-Recall Curve"
)

print(pr_plot)
```

# Saving Visualization Results

The `save_visualization_plots` function saves all visualization plots to a specified directory:

```{r save_plots, eval=FALSE}
# Create directory for results
results_dir <- "scCulturePredict_results"
if (!dir.exists(results_dir)) dir.create(results_dir)

# Save visualization plots
save_visualization_plots(
    seurat_object = seurat_object,
    evaluation_results = evaluation_results,
    output_dir = results_dir,
    prefix = "example",
    color_palette = custom_colors,
    verbose = TRUE
)
```

This will save the following files to the specified directory:
- example_umap.png
- example_direct_accuracy.png
- example_svm_accuracy.png
- example_direct_confusion.png
- example_svm_confusion.png

# Advanced Customization

Since all visualization functions return `ggplot` objects, you can further customize them using standard `ggplot2` functions:

```{r ggplot_customize}
# Start with a basic plot
base_plot <- create_umap_plot(
    seurat_object = seurat_object,
    color_by = "sample"
)

# Add custom ggplot2 elements
customized_plot <- base_plot +
    theme_minimal() +
    labs(
        title = "Highly Customized UMAP Plot",
        subtitle = "With additional annotations",
        caption = "Created with scCulturePredict"
    ) +
    theme(
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12, color = "darkgray"),
        legend.position = "bottom",
        panel.grid.major = element_line(color = "lightgray", linetype = "dashed"),
        panel.background = element_rect(fill = "aliceblue")
    ) +
    scale_color_brewer(palette = "Set1")

print(customized_plot)
```

# Conclusion

The `scCulturePredict` package provides a comprehensive set of visualization functions for exploring single-cell data and evaluating prediction results. These functions generate publication-ready figures that can be further customized using the powerful `ggplot2` framework.

# Session Info

```{r session_info}
sessionInfo()
```
