---
title: "Introduction to scCulturePredict"
author: "Niccolò Bianchi"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to scCulturePredict}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 8,
    fig.height = 6,
    warning = FALSE,
    message = FALSE
)
```

# Introduction

`scCulturePredict` is a comprehensive R package for analyzing single-cell RNA-seq data with a focus on predicting cell culture media from transcriptomic profiles. The package provides tools for loading, preprocessing, and analyzing single-cell data using pathway analysis and dimensionality reduction techniques.

This vignette provides an introduction to the core functionality of the `scCulturePredict` package and demonstrates a basic workflow using example data.

## Features

The `scCulturePredict` package offers several key features:

* Data loading and preprocessing with robust error handling
* Dimensionality reduction using UMAP
* KEGG pathway analysis for biological interpretation
* Prediction of cell culture media using similarity-based and machine learning approaches
* Evaluation and visualization of prediction results

# Installation

## From Bioconductor

The `scCulturePredict` package is available on Bioconductor and can be installed using the following commands:

```{r install, eval=FALSE}
# Package installation (for reference only - package should be installed separately):
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("scCulturePredict")
```

## From GitHub

The development version can be installed directly from GitHub:

```{r install_github, eval=FALSE}
# install.packages("devtools")
devtools::install_github("nccb/scCulturePredict")
```

# Getting Started

## Loading the Package

```{r setup}
library(scCulturePredict)
library(Seurat)
library(ggplot2)
```

## Example Data

`scCulturePredict` comes with example data in two formats that can be accessed using the `system.file()` function:

1. CSV format - Simple comma-separated files with counts matrix and metadata
2. 10X Genomics format - Industry standard format with barcodes.tsv, features.tsv, and matrix.mtx files

```{r example_data}
# Get the path to CSV format example data
csv_data_dir <- system.file("extdata", "example_data", package = "scCulturePredict")

# Get the path to 10X format example data
tenx_data_dir <- system.file("extdata", "example_data_10x", package = "scCulturePredict")
```

# Basic Workflow

The typical workflow for using `scCulturePredict` consists of the following steps:

1. Load single-cell data
2. Preprocess the data
3. Perform dimensionality reduction
4. Conduct pathway analysis
5. Predict cell culture media
6. Evaluate and visualize results

## Quick Start: Complete Analysis with One Function

For most users, the easiest way to run the complete scCulturePredict analysis is using the main `scCulture()` function, which performs all steps automatically:

```{r complete_analysis, eval=FALSE}
# Complete analysis with one function call
results <- scCulture(
    tenx_data_dir = csv_data_dir,
    input_type = "10x",
    kegg_file = system.file("extdata", "kegg", "example_pathway.keg", package = "scCulturePredict"),
    output_dir = "./scCulture_results",
    experiment_id = "example_analysis",
    mode = "build",
    progress = TRUE, # Show progress bar
    verbose = TRUE # Print detailed messages
)

# Access the final results
seurat_object <- results$seurat_object
pathway_results <- results$pathway_results
prediction_results <- results$prediction_results
evaluation_results <- results$evaluation_results

# View UMAP coordinates with predictions
head(seurat_object@meta.data[, c("UMAP_1", "UMAP_2", "sample", "classification_pred")])

# Check prediction accuracy
table(seurat_object$sample, seurat_object$classification_pred)
```

The `scCulture()` function automatically:
- Loads and validates your data
- Performs preprocessing and quality control
- Generates UMAP coordinates
- Conducts KEGG pathway analysis
- Makes predictions using both similarity and machine learning approaches
- Evaluates prediction performance
- Creates publication-ready visualizations
- Saves all results to the specified output directory

All plots and results are automatically saved to `./scCulture_results/` for immediate use!

## Advanced Usage: Step-by-Step Workflow

For users who need more control over individual steps or want to understand the underlying processes, the following sections demonstrate each step in detail.

## 1. Load Data

The `scCulturePredict` package can load data in both CSV and 10X Genomics formats. Here are examples of both approaches:

### Loading CSV Format Data

```{r load_csv_data, eval=FALSE}
# Load the CSV format example data (legacy format)
seurat_object_csv <- load_10x_data(
    tenx_data_dir = csv_data_dir,
    experiment_id = "example",
    min_cells = 3,
    min_features = 10,
    verbose = TRUE
)
```

### Loading 10X Genomics Format Data

```{r load_10x_data, eval=FALSE}
# Load the 10X format example data
seurat_object_10x <- load_10x_data(
    tenx_data_dir = tenx_data_dir,
    experiment_id = "example_10x",
    min_cells = 3,
    min_features = 10,
    verbose = TRUE
)

# Alternatively, you can use Seurat's Read10X function directly
tenx_data <- Read10X(tenx_data_dir)
seurat_object_10x_direct <- CreateSeuratObject(
    counts = tenx_data,
    min.cells = 3,
    min.features = 10
)
```

### Working with SingleCellExperiment Objects

The package also supports SingleCellExperiment objects, which are the standard data structure for Bioconductor single-cell packages:

```{r load_sce_data, eval=FALSE}
# Load SingleCellExperiment from an RDS file
seurat_object_sce <- load_sce_data(
    sce_data_path = "path/to/sce_data.rds",
    experiment_id = "example_sce",
    min_cells = 3,
    min_features = 10,
    verbose = TRUE
)

# Or use an existing SingleCellExperiment object directly
library(SingleCellExperiment)
sce <- SingleCellExperiment(
    assays = list(counts = matrix(rpois(1000, lambda = 5), nrow = 100, ncol = 10)),
    colData = DataFrame(sample = rep(c("A", "B"), 5))
)

seurat_object_from_sce <- load_sce_data(
    sce_data_path = sce,  # Can pass the object directly
    experiment_id = "example_sce",
    min_cells = 3,
    min_features = 10,
    verbose = TRUE
)
```

### Using scCulture with Different Input Types

The `scCulture()` function supports both 10X and SingleCellExperiment formats through the `input_type` parameter:

```{r scculture_input_types, eval=FALSE}
# Using 10X data
results_10x <- scCulture(
    tenx_data_dir = tenx_data_dir,
    input_type = "10x",
    kegg_file = system.file("extdata", "kegg", "example_pathway.keg", package = "scCulturePredict"),
    output_dir = "./results_10x",
    mode = "build",
    experiment_id = "10x_analysis",
    verbose = TRUE
)

# Using SingleCellExperiment data
results_sce <- scCulture(
    sce_data_path = "path/to/sce_data.rds",  # or pass an SCE object directly
    input_type = "sce",
    kegg_file = system.file("extdata", "kegg", "example_pathway.keg", package = "scCulturePredict"),
    output_dir = "./results_sce",
    mode = "build",
    experiment_id = "sce_analysis",
    verbose = TRUE
)
```

### Cross-Dataset Predictions

One powerful feature is the ability to train on one data format and predict on another:

```{r cross_dataset, eval=FALSE}
# Train model on 10X data
model_10x <- scCulture(
    tenx_data_dir = tenx_data_dir,
    input_type = "10x",
    kegg_file = system.file("extdata", "kegg", "example_pathway.keg", package = "scCulturePredict"),
    output_dir = "./model_10x",
    mode = "build",
    experiment_id = "training_10x",
    verbose = TRUE
)

# Apply the trained model to SCE data
predictions_on_sce <- scCulture(
    sce_data_path = "path/to/unlabeled_sce.rds",
    input_type = "sce",
    output_dir = "./predictions_sce",
    mode = "predict",
    fingerprint_file = model_10x$fingerprint_file,
    experiment_id = "predict_sce",
    verbose = TRUE
)
```

```{r mock_data, echo=FALSE}
# For the vignette, create a mock Seurat object since we're not evaluating
# the above code chunks
counts <- read.csv(file.path(csv_data_dir, "counts.csv"), row.names = 1)
metadata <- read.csv(file.path(csv_data_dir, "metadata.csv"), row.names = 1)
seurat_object <- CreateSeuratObject(counts = as.matrix(counts), meta.data = metadata)

# Add UMAP coordinates for visualization
set.seed(42)
seurat_object@meta.data$UMAP_1 <- rnorm(ncol(seurat_object))
seurat_object@meta.data$UMAP_2 <- rnorm(ncol(seurat_object))
```

## 2. Preprocess Data

Next, we preprocess the data, which includes normalization, scaling, and identifying variable features:

```{r preprocess, eval=FALSE}
# Preprocess the data
seurat_object <- preprocess_data(
    seurat_object = seurat_object,
    normalization_method = "LogNormalize",
    scale_factor = 10000,
    variable_features_n = 2000,
    verbose = TRUE
)
```

## 3. Perform Dimensionality Reduction

We then reduce the dimensionality of the data using UMAP:

```{r dimred, eval=FALSE}
# Perform dimensionality reduction
seurat_object <- reduce_dimensions(
    seurat_object = seurat_object,
    pca_dims = 30,
    umap_dims = 2,
    verbose = TRUE
)
```

Let's visualize the UMAP embedding:

```{r umap_vis}
# Create visualization using plot_scCulture
# Note: plot_scCulture expects the results from scCulture() function
# If you have results from scCulture, you can visualize them:
# umap_plot <- plot_scCulture(results)

# For direct visualization when UMAP coordinates are in metadata:
library(ggplot2)
if(all(c("UMAP_1", "UMAP_2") %in% colnames(seurat_object@meta.data))) {
  ggplot(seurat_object@meta.data, aes(x = UMAP_1, y = UMAP_2, color = sample)) +
    geom_point(size = 1.5) +
    theme_minimal() +
    labs(title = "UMAP of Example Data")
} else {
  message("UMAP coordinates not found in metadata")
}
```

## 4. Pathway Analysis

Now we perform KEGG pathway analysis to identify biological pathways that are active in our data:

```{r pathway, eval=FALSE}
# Path to KEGG data
kegg_file <- system.file("extdata", "kegg", "example_pathway.keg", package = "scCulturePredict")

# Parse KEGG data
kegg_data <- parse_kegg_keg(kegg_file, verbose = TRUE)

# Build pathway fingerprints
pathway_matrix <- build_fingerprints(
    seurat_object = seurat_object,
    kegg_data = kegg_data,
    verbose = TRUE
)
```

```{r mock_pathway, echo=FALSE}
# Create mock pathway data for the vignette
set.seed(42)
pathway_matrix <- matrix(rnorm(30 * 10), nrow = 30, ncol = 10)
rownames(pathway_matrix) <- paste0("Pathway", 1:30)
colnames(pathway_matrix) <- paste0("Cell", 1:10)

# Create mock signature matrix
signature_matrix <- matrix(rnorm(10 * 3), nrow = 10, ncol = 3)
rownames(signature_matrix) <- paste0("Pathway", 1:10)
colnames(signature_matrix) <- c("A", "B", "C")
```

## 5. Predict Cell Culture Media

We can now predict the culture media using both similarity-based and SVM approaches:

```{r predict, eval=FALSE}
# Similarity-based prediction
similarity_results <- predict_by_similarity(
    pathway_matrix = pathway_matrix,
    signature_matrix = signature_matrix,
    threshold = 0.1
)

# SVM-based prediction
svm_results <- predict_by_svm(
    pathway_matrix = pathway_matrix,
    seurat_object = seurat_object
)
```

```{r mock_predict, echo=FALSE}
# Create mock prediction results
similarity_results <- list(
    similarity_matrix = matrix(runif(90), nrow = 30, ncol = 3),
    predicted_direct = sample(c("A", "B", "C"), 30, replace = TRUE),
    predicted_threshold = sample(c("A", "B", "C", NA), 30, replace = TRUE)
)

svm_results <- list(
    predictions = factor(sample(c("A", "B", "C"), 30, replace = TRUE)),
    accuracy = 0.85,
    confusion_matrix = matrix(c(8, 1, 1, 1, 7, 2, 0, 1, 9),
        nrow = 3,
        dimnames = list(c("A", "B", "C"), c("A", "B", "C"))
    )
)
```

## 6. Evaluate and Visualize Results

First, let's add some mock prediction results to demonstrate the evaluation functionality:

```{r mock_predictions, echo=FALSE}
# Add mock prediction columns to demonstrate evaluation
set.seed(42)
n_cells <- ncol(seurat_object)

# Create mock predictions with some realistic accuracy
seurat_object$predicted_sample_1 <- sample(c("A", "B", "C"), n_cells, replace = TRUE, prob = c(0.4, 0.35, 0.25))
seurat_object$predicted_sample_2 <- sample(c("A", "B", "C"), n_cells, replace = TRUE, prob = c(0.35, 0.4, 0.25))
seurat_object$classification_pred <- sample(c("A", "B", "C"), n_cells, replace = TRUE, prob = c(0.35, 0.35, 0.3))

# Make predictions somewhat correlated with actual samples for realism
actual_samples <- seurat_object$sample
for (i in 1:n_cells) {
    if (runif(1) < 0.7) { # 70% accuracy for demonstration
        seurat_object$predicted_sample_1[i] <- actual_samples[i]
        seurat_object$predicted_sample_2[i] <- actual_samples[i]
        seurat_object$classification_pred[i] <- actual_samples[i]
    }
}
```

Finally, we evaluate the prediction results and create visualizations:

```{r evaluate}
# Evaluate predictions
evaluation_results <- evaluate_predictions(seurat_object)
```

# Create evaluation plots
# Note: Individual plotting functions have been moved to inst/extras/alternative_implementations.R
# Use create_evaluation_plots() for comprehensive visualization
evaluation_plots <- create_evaluation_plots(
  seurat_object = seurat_object,
  results_dir = "./results"
)

# Display the plots
print(evaluation_plots)

# For detailed evaluation metrics visualization
# The create_evaluation_metrics_plot() function has been moved to inst/extras/alternative_implementations.R
# metrics_plot <- create_evaluation_metrics_plot(
#   evaluation_results = evaluation_results,
#   plot_type = "confusion",
#   title = "Prediction Results"
# )
#
# # Display the plot
# print(metrics_plot)
```

# Saving Results

You can save all visualization plots to a directory:

```{r save, eval=FALSE}
# Create directory for results
results_dir <- "scCulturePredict_results"
if (!dir.exists(results_dir)) dir.create(results_dir)

# Save plots using ggplot2::ggsave or base R functions
# Note: save_visualization_plots has been moved to inst/extras/alternative_implementations.R

# Save evaluation plots created earlier
if (exists("evaluation_plots")) {
  for (i in seq_along(evaluation_plots)) {
    filename <- file.path(results_dir, paste0("example_plot_", i, ".png"))
    ggplot2::ggsave(filename, evaluation_plots[[i]], width = 10, height = 8)
  }
}

# Save metrics plot
if (exists("metrics_plot")) {
  ggplot2::ggsave(file.path(results_dir, "example_metrics.png"),
                  metrics_plot, width = 10, height = 8)
}
```

# Working with Different Data Formats

## CSV Format

The CSV format is simpler and consists of:
- A counts matrix with genes as rows and cells as columns
- A metadata file with cells as rows and annotations as columns

This format is easy to create and manipulate using standard R functions like `read.csv()` and `write.csv()`.

## 10X Genomics Format

The 10X Genomics format is an industry standard and consists of:
- `barcodes.tsv` - Contains cell barcodes
- `features.tsv` - Contains gene/feature information
- `matrix.mtx` - Contains the count matrix in sparse format

This format is more memory-efficient for large datasets and is widely used in single-cell analysis.

# Conclusion

This vignette has demonstrated the basic workflow for using the `scCulturePredict` package to analyze single-cell RNA-seq data and predict cell culture media, with support for both CSV and 10X Genomics data formats. For more advanced usage and customization options, please refer to the advanced vignette and the function documentation.

# Session Info

```{r session_info}
sessionInfo()
```
