---
title: "Visualization with scCulturePredict"
author: "Niccolò Bianchi"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Visualization with scCulturePredict}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 8,
    fig.height = 6,
    warning = FALSE,
    message = FALSE
)
```

# Introduction

Effective visualization is crucial for interpreting single-cell RNA sequencing data. The `scCulturePredict` package provides a comprehensive set of visualization functions designed to help you explore your data and evaluate prediction results. This vignette demonstrates how to use these visualization tools to create informative and publication-ready figures.

## Loading Required Packages

```{r setup}
library(scCulturePredict)
library(Seurat)
library(ggplot2)
library(patchwork)
```

## Creating Example Data

For this vignette, we'll create some example data to demonstrate the visualization capabilities:

```{r example_data}
# Create a mock Seurat object
create_mock_seurat <- function() {
    # Create a small Seurat object with UMAP coordinates
    set.seed(42)
    counts <- matrix(rpois(100 * 30, 10), nrow = 100, ncol = 30)
    rownames(counts) <- paste0("gene", 1:100)
    colnames(counts) <- paste0("cell", 1:30)

    seurat <- Seurat::CreateSeuratObject(counts = counts)

    # Add UMAP coordinates and metadata
    seurat@meta.data$UMAP_1 <- rnorm(30)
    seurat@meta.data$UMAP_2 <- rnorm(30)
    seurat@meta.data$sample <- rep(c("A", "B", "C"), each = 10)
    seurat@meta.data$cluster <- rep(c(1, 2, 3), times = 10)
    seurat@meta.data$percent_mito <- runif(30, 0, 10)

    return(seurat)
}

# Create mock evaluation results
create_mock_evaluation <- function() {
    list(
        direct_accuracy = data.frame(
            sample = c("A", "B", "C"),
            correct = c(8, 7, 9),
            percent = c("80%", "70%", "90%")
        ),
        direct_table = as.table(matrix(c(8, 1, 1, 1, 7, 2, 1, 2, 7),
            nrow = 3,
            dimnames = list(c("A", "B", "C"), c("A", "B", "C"))
        )),
        svm_accuracy = data.frame(
            sample = c("A", "B", "C"),
            correct = c(9, 8, 7),
            percent = c("90%", "80%", "70%")
        ),
        svm_table = as.table(matrix(c(9, 0, 1, 1, 8, 1, 0, 2, 7),
            nrow = 3,
            dimnames = list(c("A", "B", "C"), c("A", "B", "C"))
        )),
        threshold_table = as.table(matrix(c(7, 1, 0, 2, 6, 1, 1, 2, 8),
            nrow = 3,
            dimnames = list(c("A", "B", "C"), c("A", "B", "C"))
        ))
    )
}

# Generate example data
seurat_object <- create_mock_seurat()
evaluation_results <- create_mock_evaluation()
```

# Basic Visualizations

## UMAP Visualization

For UMAP visualizations, use Seurat's built-in `DimPlot` function or `plot_scCulture` for scCulture results:

```{r umap_basic}
# Basic UMAP plot using metadata coordinates
library(ggplot2)
umap_plot <- ggplot(seurat_object@meta.data,
                    aes(x = UMAP_1, y = UMAP_2, color = sample)) +
    geom_point(size = 2) +
    theme_minimal() +
    labs(title = "UMAP of Example Data",
         x = "UMAP_1",
         y = "UMAP_2",
         color = "Sample") +
    scale_color_manual(values = c("A" = "#E41A1C", "B" = "#377EB8", "C" = "#4DAF4A"))

print(umap_plot)
```

### Customizing UMAP Plots

You can customize various aspects of the UMAP plot:

```{r umap_custom}
# Customized UMAP plot
umap_custom <- ggplot(seurat_object@meta.data,
                      aes(x = UMAP_1, y = UMAP_2, color = factor(cluster))) +
    geom_point(size = 2.5) +
    theme_minimal() +
    labs(title = "UMAP by Cluster",
         x = "UMAP_1",
         y = "UMAP_2",
         color = "Cluster ID") +
    scale_color_discrete()

print(umap_custom)
```

You can also use other metadata columns for coloring:

```{r umap_continuous}
# UMAP with continuous variable
umap_continuous <- ggplot(seurat_object@meta.data,
                          aes(x = UMAP_1, y = UMAP_2, color = percent_mito)) +
    geom_point(size = 2) +
    theme_minimal() +
    labs(title = "UMAP by Mitochondrial Percentage",
         x = "UMAP_1",
         y = "UMAP_2",
         color = "% Mito") +
    scale_color_viridis_c()

print(umap_continuous)
```

### Combining Multiple UMAP Plots

The plots can be combined using the `patchwork` package:

```{r umap_combined}
# Combine multiple UMAP plots
combined_plot <- umap_plot + umap_custom + umap_continuous +
    plot_layout(ncol = 2) +
    plot_annotation(title = "Multiple UMAP Visualizations")

print(combined_plot)
```

# Prediction Evaluation Visualizations

## Evaluation Metrics Plots

Note: The `create_evaluation_metrics_plot` function has been moved to inst/extras/alternative_implementations.R

```{r accuracy}
# Metrics visualization
# The create_evaluation_metrics_plot function is now in inst/extras/alternative_implementations.R
# metrics_plot <- create_evaluation_metrics_plot(
#     evaluation_results = evaluation_results,
#     plot_type = "metrics",
#     title = "Direct Prediction Accuracy"
# )

# Alternative: Use create_evaluation_plots for comprehensive visualization
evaluation_plots <- create_evaluation_plots(
    seurat_object = seurat_object,
    results_dir = "./results",
    verbose = FALSE
)

# Display the plots (plots are saved to results directory)
# evaluation_plots
```

### Customizing Accuracy Plots

You can customize the color palette:

```{r accuracy_custom}
# Custom color palette
custom_colors <- c("A" = "#E41A1C", "B" = "#377EB8", "C" = "#4DAF4A")

# Customized accuracy plot
# The create_evaluation_metrics_plot function is now in inst/extras/alternative_implementations.R
# custom_accuracy <- create_evaluation_metrics_plot(
#     evaluation_results = evaluation_results,
#     plot_type = "metrics",
#     title = "Custom Accuracy Plot",
#     color_palette = c("#2E86AB", "#A23B72", "#F18F01")
# )
#
# print(custom_accuracy)
```

## Confusion Matrix Visualization

Note: The `create_evaluation_metrics_plot` function has been moved to inst/extras/alternative_implementations.R

```{r confusion}
# Confusion matrix visualization
# The create_evaluation_metrics_plot function is now in inst/extras/alternative_implementations.R
# confusion_plot <- create_evaluation_metrics_plot(
#     evaluation_results = evaluation_results,
#     plot_type = "confusion",
#     title = "Direct Prediction Confusion Matrix"
# )

# Alternative confusion matrix view
# Note: Individual confusion matrix functions have been moved to inst/extras/alternative_implementations.R
# svm_confusion <- create_evaluation_metrics_plot(
#     evaluation_results = evaluation_results,
#     plot_type = "confusion",
#     title = "SVM Prediction Confusion Matrix"
# )

# For comprehensive evaluation including all methods
# Use create_evaluation_plots which generates multiple visualizations
all_plots <- create_evaluation_plots(
    seurat_object = seurat_object,
    results_dir = "./results",
    verbose = FALSE
)

# Individual plots can be accessed from the results
# Display the plots
# Note: Individual plot objects would need to be created separately
# direct_confusion
# svm_confusion
# threshold_confusion
```

### Customizing Confusion Heatmaps

You can customize the color gradient:

```{r confusion_custom}
# Custom color gradient
custom_gradient <- c("white", "navy")

# Customized confusion matrix visualization
# The create_evaluation_metrics_plot function is now in inst/extras/alternative_implementations.R
# custom_confusion <- create_evaluation_metrics_plot(
#     evaluation_results = evaluation_results,
#     plot_type = "confusion",
#     color_palette = c("#2E86AB", "#A23B72", "#F18F01"),
#     title = "Custom Confusion Matrix"
# )
#
# print(custom_confusion)
```

# Evaluation Metrics Visualization

Note: The `create_evaluation_metrics_plot` function has been moved to inst/extras/alternative_implementations.R

```{r eval_metrics, eval=FALSE}
# For this demonstration, we need additional metrics that would come from a real evaluation
# This code is provided as an example but not evaluated in this vignette

# Create evaluation metrics plot
# The create_evaluation_metrics_plot function is now in inst/extras/alternative_implementations.R
# metrics_plot <- create_evaluation_metrics_plot(
#     evaluation_results = complete_evaluation_results,
#     plot_type = "metrics",
#     title = "Prediction Performance Metrics"
# )
#
# print(metrics_plot)
#
# # ROC curve
# roc_plot <- create_evaluation_metrics_plot(
#     evaluation_results = complete_evaluation_results,
#     plot_type = "roc",
#     title = "ROC Curve"
# )
#
# print(roc_plot)
#
# # Precision-Recall curve
# pr_plot <- create_evaluation_metrics_plot(
#     evaluation_results = complete_evaluation_results,
#     plot_type = "pr",
#     title = "Precision-Recall Curve"
# )
#
# print(pr_plot)
```

# Saving Visualization Results

Visualization plots can be saved using `ggplot2::ggsave()` or base R graphics functions:

```{r save_plots, eval=FALSE}
# Create directory for results
results_dir <- "scCulturePredict_results"
if (!dir.exists(results_dir)) dir.create(results_dir)

# Save individual plots using ggplot2::ggsave
# Note: save_visualization_plots has been moved to inst/extras/alternative_implementations.R

# Save confusion matrix plot
if (exists("confusion_plot")) {
  ggplot2::ggsave(
    filename = file.path(results_dir, "example_confusion.png"),
    plot = confusion_plot,
    width = 10,
    height = 8,
    dpi = 300
  )
}

# Save metrics plot
if (exists("metrics_plot")) {
  ggplot2::ggsave(
    filename = file.path(results_dir, "example_metrics.png"),
    plot = metrics_plot,
    width = 10,
    height = 8,
    dpi = 300
  )
}

# Save UMAP plot
if (exists("umap_plot")) {
  ggplot2::ggsave(
    filename = file.path(results_dir, "example_umap.png"),
    plot = umap_plot,
    width = 10,
    height = 8,
    dpi = 300
  )
}
```

This will save the following files to the specified directory:
- example_umap.png
- umap_experimental.png (from create_evaluation_plots)

# Advanced Customization

Since all visualization functions return `ggplot` objects, you can further customize them using standard `ggplot2` functions:

```{r ggplot_customize}
# Create a highly customized plot using ggplot2 directly
customized_plot <- ggplot(seurat_object@meta.data,
                          aes(x = UMAP_1, y = UMAP_2, color = sample)) +
    geom_point(size = 2) +
    theme_minimal() +
    labs(
        title = "Highly Customized UMAP Plot",
        subtitle = "With additional annotations",
        caption = "Created with scCulturePredict",
        x = "UMAP_1",
        y = "UMAP_2",
        color = "Sample"
    ) +
    theme(
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12, color = "darkgray"),
        legend.position = "bottom",
        panel.grid.major = element_line(color = "lightgray", linetype = "dashed"),
        panel.background = element_rect(fill = "aliceblue")
    ) +
    scale_color_brewer(palette = "Set1")

print(customized_plot)
```

# Conclusion

The `scCulturePredict` package provides a comprehensive set of visualization functions for exploring single-cell data and evaluating prediction results. These functions generate publication-ready figures that can be further customized using the powerful `ggplot2` framework.

# Session Info

```{r session_info}
sessionInfo()
```
